---
title: "Computing the substitution matrix for CDR3 amino acid sequence matching"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data and general stats

Libraries & load data

```{r}
set.seed(51103)

library(MASS)
library(dplyr)
library(pROC)
library(ggplot2)
library(reshape2)
library(ggbeeswarm)
library(gplots)
library(RColorBrewer)
select = dplyr::select

df = read.table("mutations.txt.gz", header=T, sep="\t") %>%
  select(same.ag, align.id, mut.type, mut.from, mut.to, mut.region, gene)
df$same.ag = df$same.ag > 0

head(df)

summary(df)
```

```{r}
n = 50
df.bs = sample_n(df, nrow(df) * n, replace=T)
df.bs$bs = rep(1:n, each = nrow(df))
```

Number of substitutions for same/different antigen by region

```{r}
df.bs$match = df.bs$mut.from == df.bs$mut.to

df.summary = df.bs %>%
  group_by(bs, align.id, mut.region, same.ag) %>%
  summarize(match = sum(match), mismatch = n() - match)
```

```{r}
df.summary$mut.region <- factor(df.summary$mut.region, levels = c("V", "N", "J"))

ggplot(df.summary, aes(x = mismatch, group = interaction(bs,same.ag), color=same.ag)) +
  geom_freqpoly(aes(y=..density..), binwidth = 1, alpha=0.3) +
  facet_grid(. ~ mut.region) +
  scale_x_continuous(limits=c(0, 3), breaks=0:3) +
  scale_color_brewer(palette = "Set1") +
  theme_bw()
```

### Substitution matrix

Computing substitution matrix for bootstrapped samples

```{r}
compute_prob = function(mut.from, mut.to, count) {
  .df = data.frame(mut.from, mut.to, count)
  
  .mat = dcast(.df, 
                mut.from ~ mut.to, value.var = "count", 
                fun.aggregate = sum)
  rownames(.mat) = .mat$mut.from
  .mat$mut.from = NULL
  .mat = as.matrix(.mat)
  
  rs = rowSums(.mat)
  p0 = diag(.mat) / rs
  
  .mat = .mat / (rs - diag(.mat)) * p0 # subst prob
  diag(.mat) = p0 #match prob
  
  .mat = (.mat ^ 2 + t(.mat) ^ 2) / (.mat + t(.mat)) # symmetrize #2
  
  merge(.df, melt(.mat,varnames = c("mut.from", "mut.to")))$value
}

df.ss = df.bs
df.ss$mut.from = df.bs$mut.to
df.ss$mut.to = df.bs$mut.from
df.ss = rbind(df.bs, df.ss)

df.ss2 = df.ss
df.ss2$mut.region <- "Any"

df.ss = rbind(df.ss, df.ss2)

df.ss = df.ss %>% 
  group_by(bs, same.ag, mut.region, mut.from, mut.to) %>% 
  summarise(count = n()) %>%
  droplevels() %>%
  group_by(bs, same.ag, mut.region) %>% 
  mutate(P = compute_prob(mut.from, mut.to, count)) %>%
  group_by(bs, mut.region, mut.from, mut.to) %>% 
  summarise(Q = ifelse(n() == 2, log10(P[same.ag] / P[!same.ag]), NA))

df.ss$mut.region <- factor(df.ss$mut.region, levels = c("V", "N", "J", "Any"))

head(df.ss)
```

Substitution matrix. Bootstrapped means are plotted for matrices obtained for selected regions and all regions. Dots indicate bootstrap value Z-score > 1.96 threshold compared to Q = 0.

```{r}
df.ss.m = df.ss %>%
  group_by(mut.region, mut.from, mut.to) %>% 
  summarise(Q.sd = sd(Q), Q = mean(Q))

write.table(df.ss.m, "substitution.matrix.txt", sep = "\t", quote = F, row.names = F)

ggplot(df.ss.m, aes(x=mut.from, y=mut.to, fill = Q)) + 
  geom_tile() + 
  geom_point(shape = 21, aes(color = abs(Q / Q.sd) > 1.96)) +
  facet_grid(~mut.region) +
  scale_fill_gradient2(midpoint = 0, low = "#5e4fa2", mid = "#ffffbf", high = "#9e0142", 
                       na.value = "white", limits = c(-1.5,1.5)) + 
  scale_color_manual(values = c(NA, "black", "NA"), guide=F) +
  theme_bw() + theme(panel.background = element_rect(fill = "white"),
                     panel.grid.major = element_blank(), panel.grid.minor = element_blank())

ggplot(subset(df.ss.m, mut.region == "Any"), aes(x=mut.from, y=mut.to, fill = Q)) + 
  geom_tile() + 
  geom_point(shape = 21, aes(color = abs(Q / Q.sd) > 1.96)) +
  scale_fill_gradient2(midpoint = 0, low = "#5e4fa2", mid = "#ffffbf", high = "#9e0142", 
                       na.value = "white", limits = c(-1.5,1.5)) + 
  scale_color_manual(values = c(NA, "black", "NA"), guide=F) +
  theme_bw() + theme(panel.background = element_rect(fill = "white"),
                     panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

Checking that we don't have high Q scores with low Z-score (indicating anecdotal cases with rare amino acids)

```{r}
df.ss.m$replacement = ifelse(df.ss.m$mut.from == df.ss.m$mut.to, "match", "mismatch")

ggplot(df.ss.m, aes(x=Q, y = abs(Q / Q.sd), color = abs(Q / Q.sd) > 1.96)) + 
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray30") +
  geom_point(shape = 21) +
  geom_errorbarh(aes(xmin = Q - Q.sd, xmax = Q + Q.sd)) +
  facet_grid(replacement~mut.region) +
  scale_x_continuous(limits=c(-1.5, 1.5)) +
  ylab("Z") +
  scale_color_manual("Z>1.96", values = c("black", "red", "black")) +
  theme_bw()
```

Replacement-conservation score contrast.

```{r}
ggplot(df.ss.m, aes(x = Q, group=mut.region, color=mut.region)) + 
  stat_ecdf(aes(linetype = mut.region == "Any")) + 
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(limits=c(-2,2))+
  facet_wrap(~replacement) +
  theme_bw()
```

Cluster amino acids based on substitution matrix

```{r}
df.hydro <- data.frame(
  aa = strsplit("I V L F C M A W G T S Y P H N D Q E K R", " ")[[1]],
  hydrop = strsplit("4.5 4.2 3.8 2.8 2.5 1.9 1.8 -0.9 -0.4 -0.7 -0.8 -1.3 -1.6 -3.2 -3.5 -3.5 -3.5 -3.5 -3.9 -4.5", " ")[[1]]
)

df.hydro = df.hydro %>%
  mutate(hydrop = as.numeric(as.character(hydrop))) %>%
  arrange(hydrop) %>%
  mutate(hydrop.sc = round(100 * (hydrop - min(hydrop)) / (max(hydrop) - min(hydrop))))

df.hydro$color = colorRampPalette(brewer.pal(11, 'PRGn'))(101)[df.hydro$hydrop.sc + 1]

aa_colors = df.hydro$color
names(aa_colors) = df.hydro$aa

###

mat.ss.m = subset(df.ss.m, mut.region == "Any" & mut.from != "C" & mut.to != "C") %>%
  select(mut.from, mut.to, Q)

mat.ss.m = dcast(as.data.frame(mat.ss.m), mut.from ~ mut.to)
mat.ss.m[is.na(mat.ss.m)] = 0

rownames(mat.ss.m) = mat.ss.m$mut.from
mat.ss.m$mut.from = NULL

mat.ss.m = as.matrix(mat.ss.m)

heatmap.2(mat.ss.m, scale="none", symm = T, trace = "none",
          hclustfun = function(x) hclust(x, method = "ward"),
          col=colorRampPalette(rev(brewer.pal(11, 'RdBu')))(100),
          RowSideColors = aa_colors[rownames(mat.ss.m)], 
          ColSideColors = aa_colors[colnames(mat.ss.m)])

mds = isoMDS(as.dist(10^(-mat.ss.m)))
plot(mds$points, type = "n") #
text(mds$points, labels = rownames(mat.ss.m))
```

Strong amino acids are less likely to be replaced

```{r}
df.ss.aa = subset(df.ss.m, mut.from == mut.to)

# Useful faceted axis reordering magick
df.ss.aa$strong = df.ss.aa$mut.from %in% c("F", "I", "L", "M", "V", "W", "Y")
df.ss.aa$lbl.ds = paste(df.ss.aa$mut.from, df.ss.aa$mut.region, sep = ".")
df.ss.aa$lbl.ds = reorder(df.ss.aa$lbl.ds, df.ss.aa$Q)

ggplot(df.ss.aa, aes(x = lbl.ds, y = Q, fill=strong)) + 
  geom_bar(stat="identity") + 
  scale_fill_manual(values = c("black", "red")) +
  facet_wrap(~mut.region, scales="free_x") +
  scale_x_discrete(labels=function(x) sapply(strsplit(x,"[.]"),"[",1)) +
  theme_bw()

summary(aov(Q ~ mut.region + strong, df.ss.aa))
```

### Comparison with BLOSUM62

```{r}
df.bl62 = read.table("blosum62.txt", header = T)

df.bl62 = subset(df.bl62, tmp != "*" & tmp != "X")
rownames(df.bl62) = df.bl62[,1]
df.bl62[,1] = NULL
df.bl62[,(ncol(df.bl62)-1):ncol(df.bl62)]  <- NULL

df.bl62 = data.frame(mut.from=rep(row.names(df.bl62),ncol(df.bl62)),
                mut.to=rep(colnames(df.bl62),each=nrow(df.bl62)),
                score=unlist(c(df.bl62)))

df.bl62 = merge(subset(df.ss.m, mut.region == "Any"), df.bl62)
df.bl62$type = with(df.bl62, ifelse(mut.from == mut.to, "match", "mismatch"))
df.bl62 = subset(df.bl62, as.integer(mut.from) <= as.integer(mut.to))
df.bl62$Q[is.na(df.bl62$Q)] = 0

ggplot(df.bl62, aes(x = score, group = score, y = Q)) +
  geom_boxplot() +
  geom_beeswarm() +
  ylab("Q score") +
  scale_x_continuous("BLOSUM62 score", breaks=-20:20) +
  facet_wrap(~type, scales = "free_x") + 
  theme_bw()

summary(aov(Q ~ score, df.bl62))
```

### Classifier performance

Aux 

```{r}
df.scoring = merge(df, df.ss.m) %>%
  group_by(align.id, gene, same.ag) %>%
  summarise(score = sum(Q, na.rm=T))

df.scoring$matrix.type = "by-region"

df.scoring2 = df
df.scoring2$mut.region = NULL
df.ss2 = subset(df.ss.m, mut.region == "Any")

df.scoring2 = merge(df.scoring2, df.ss.m) %>%
  group_by(align.id, gene, same.ag) %>%
  summarise(score = sum(Q, na.rm=T))

df.scoring2$matrix.type = "all-regions"

df.scoring = rbind(df.scoring, df.scoring2)

# Compute ROCs

df.scoring.roc = data.frame()

roc2df = function(gene, mattype, rocobj) {
  print(gene)
  print(mattype)
  print(rocobj)
  
  with(rocobj, data.frame(gene = gene,
                          mattype = mattype,
                          sens = sensitivities,
                          spec = specificities,
                          auc = auc[1]))
}

for (i in unique(df.scoring$gene)) {
  for (j in unique(df.scoring$matrix.type)) {
    df.scoring.roc = rbind(df.scoring.roc,
                          roc2df(i, j, roc(same.ag ~ score, subset(df.scoring, gene == i & matrix.type == j))))
  }
}

df.scoring.roc = rbind(df.scoring.roc,
                        roc2df("Any", "by-region", roc(same.ag ~ score, subset(df.scoring, matrix.type == "by-region"))))
df.scoring.roc = rbind(df.scoring.roc,
                        roc2df("Any", "all-regions", roc(same.ag ~ score, subset(df.scoring, matrix.type == "all-regions"))))
```

No difference between by-region and all-region matrices, except for TRA

```{r}
ggplot(df.scoring.roc, aes(x=spec, y=sens)) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed") +
  geom_line(color="red") +
  facet_grid(mattype~gene) +
  scale_x_reverse() +
  theme_bw()
```

> TODO: show blosum not works here

